---
output:
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# DuckGeneSnap

DuckGeneSnap is a pure static, browser-side reproduction of the core GeneSnap
idea: parse personal genotype/variant files, match clinically or biologically
annotated loci, and present categorized results without running a backend.

It follows the DuckBedQC pattern: the page loads `webR`, `duckdb`, and
`Rduckhts` in the browser. User files are staged into the browser filesystem and
queried locally. No genotype data is uploaded to a server.

> **Disclaimer:** DuckGeneSnap is informational only and is not medical advice.
> Clinically significant findings require clinical-grade confirmation and review
> by a qualified professional.

## What works now

- Static `index.html`; no FastAPI/SQLite service.
- 23andMe-style raw text upload.
- VCF/BCF upload through `Rduckhts::rduckhts_bcf(..., tidy_format = TRUE)`.
- Locus-first matching: `build + normalized chromosome + 1-based position`.
- Static DuckDB annotation file staged into webR and attached read-only.
- Sorted ZSTD Parquet sidecars with configurable row group size and compression
  level.
- Auxiliary `variant_keys` table for allele-aware display, QA, and future
  exact-key refinement.
- Optional VCF/BCF liftover hook using `bcftools_liftover()` when chain and
  FASTA files are supplied by the user.
- HTMX is included for static fragment loading; computation remains in the local
  DuckDB/webR layer.
- Demo 23andMe-style file at `public/demo/example_23andme.txt`.

The committed asset bundle is intentionally small: 16 demo curated loci and 32
auxiliary VariantKey rows. Treat it as a scaffold for production asset builds.

## Run locally

Serve the repository over HTTP; do not open `index.html` with `file://`.

```bash
cd DuckGeneSnap
python3 -m http.server 8000
# open http://localhost:8000/
```

First browser use can take a while because webR installs `DBI`, `duckdb`,
`jsonlite`, and `Rduckhts` into the browser runtime.

## Rebuild static assets

Requires R packages `optparse`, `DBI`, `duckdb`, `Rduckhts`, and `jsonlite`.

```bash
cd DuckGeneSnap
Rscript scripts/build_assets.R
```

Useful size-test examples:

```bash
Rscript scripts/build_assets.R \
  --out-dir .size_tests/clinvar_hg38_noidx \
  --clinvar-grch38-bcf path/to/clinvar_hg38_p_lp.bcf \
  --skip-seed \
  --skip-indexes

Rscript scripts/build_assets.R \
  --out-dir .size_tests/gwas_full \
  --gwas-zip path/to/gwas-catalog-associations-full.zip \
  --skip-seed \
  --skip-indexes
```

Outputs:

- `public/data/duckgenesnap.duckdb`
- `public/data/variant_annotations.parquet`
- `public/data/genotype_interpretations.parquet`
- `public/data/variant_keys.parquet`
- `public/data/source_summary.parquet`
- `public/data/index_summary.parquet`
- `public/data/size_report.tsv`
- `public/data/manifest.json`

See `docs/SIZE_TESTS.md` for measured ClinVar/GWAS bundle sizes, index overhead,
ZSTD compression-level tests, `parquet-linter` output, and Parquet row group
trade-offs.

## Committed demo asset sizes

```{r asset-sizes, echo=FALSE, results='asis'}
size_path <- file.path("public", "data", "size_report.tsv")
if (file.exists(size_path)) {
  sizes <- utils::read.delim(size_path, check.names = FALSE)
  knitr::kable(sizes, align = c("l", "r", "r"))
} else {
  cat("`public/data/size_report.tsv` has not been generated yet.\n")
}
```

## Repository layout

```text
index.html                         static app shell
src/duckgenesnap.js                browser webR/DuckDB/Rduckhts runtime
public/data/duckgenesnap.duckdb    attachable annotation database
public/data/*.parquet              sorted sidecars/fallback assets
public/demo/example_23andme.txt    synthetic demo input
data/seed/*.tsv                    small demo curation source
scripts/build_assets.R             optparse asset builder/injector
docs/                              design, schema, size notes
```

## Matching and storage policy

Core ingestion does **not** depend on rsID identity. Uploaded rows are matched by:

```text
analysis_build + chrom_norm + pos
```

`source_id` may contain rsIDs or source-specific identifiers, but it is display
metadata only. `variant_keys` is kept as an auxiliary allele-specific table for
VCF/BCF QA, detail panels, and future exact allele refinement.

Do not lift or mutate VariantKeys directly. For allele-aware paths:

```text
source chrom/pos/ref/alt
  -> optional normalization/liftover
  -> recompute variantkey(chrom, pos, ref, alt)
  -> optional exact-key refinement
```

## Frontend query framework credit

The planned variant detail panels take inspiration from the lightweight
frontend-only query orchestration in Sasha Gusev's GWAS Lookup:
<https://github.com/sashagusev/gwas_lookup>. DuckGeneSnap keeps that spirit
(client-side modules, URL-stateable variant queries, parallel enrichment panels)
but uses local DuckDB/Rduckhts assets first and calls external APIs only for
on-demand detail enrichment.

## Related projects

- DuckBedQC: <https://github.com/sounkou-bioinfo/DuckBedQC>
- GeneSnap inspiration: <https://github.com/syao13/GeneSnap>
- GWAS Lookup frontend inspiration: <https://github.com/sashagusev/gwas_lookup>
- Rduckhts / duckhts: <https://github.com/RGenomicsETL/duckhts>
