---
output:
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# DuckGeneSnap

DuckGeneSnap is a pure static, browser-side reproduction of the core GeneSnap
idea: parse personal genotype/variant files, match clinically or biologically
annotated loci, and present categorized results without running a backend.

It follows the DuckBedQC pattern: the page loads `webR`, `duckdb`, and
`Rduckhts` in the browser. User files are staged into the browser filesystem and
queried locally. No genotype data is uploaded to a server.

> **Disclaimer:** DuckGeneSnap is informational only and is not medical advice.
> Clinically significant findings require clinical-grade confirmation and review
> by a qualified professional.

## What works now

- Static `index.html` with local browser-side analysis.
- 23andMe-style raw text upload.
- VCF/BCF upload through `Rduckhts::rduckhts_bcf(..., tidy_format = TRUE)`.
- Locus-first matching: `build + normalized chromosome + 1-based position`.
- Sorted ZSTD Parquet annotation assets staged into webR and queried with DuckDB.
- Configurable Parquet row group size and compression level.
- Auxiliary `variant_keys` table for allele-aware display, QA, and future
  exact-key refinement.
- Optional VCF/BCF liftover hook using `bcftools_liftover()` when chain and
  FASTA files are supplied by the user.
- HTMX is included for static fragment loading; computation remains in the local
  DuckDB/webR layer.
- Demo 23andMe-style file at `public/demo/example_23andme.txt`.

The committed annotation assets include ClinVar variant-summary rows for GRCh37
and GRCh38 (pathogenic/likely pathogenic, drug response, risk factor, and
association records), GWAS Catalog associations filtered at genome-wide
significance, and a small GRCh37 seed set used by the local demo files.

## Hosted site

GitHub Pages is configured for the repository root:

<https://sounkou-bioinfo.github.io/DuckGeneSnap/>

## Run locally

Serve the repository over HTTP; do not open `index.html` with `file://`.
For local testing, prefer goServeR because it supports range requests out of the
box:

```bash
cd DuckGeneSnap
Rscript -e "goserveR::runServer(dir='.', prefix='/', addr='127.0.0.1:8000')"
# open http://127.0.0.1:8000/
```

If needed, install goServeR from <https://github.com/sounkou-bioinfo/goServeR>.

First browser use can take a while because webR installs `DBI`, `duckdb`,
`jsonlite`, and `Rduckhts` into the browser runtime.

## Rebuild static assets

Requires R packages `optparse`, `DBI`, `duckdb`, `Rduckhts`, and `jsonlite`.

```bash
cd DuckGeneSnap
Rscript scripts/build_assets.R
```

Useful size-test examples:

```bash
Rscript scripts/build_assets.R \
  --out-dir .size_tests/clinvar_tsv_noidx \
  --clinvar-tsv path/to/variant_summary.txt.gz \
  --skip-seed \
  --skip-indexes

Rscript scripts/build_assets.R \
  --out-dir .size_tests/gwas_full \
  --gwas-zip path/to/gwas-catalog-associations-full.zip \
  --skip-seed \
  --skip-indexes
```

Outputs:

- `public/data/variant_annotations.parquet`
- `public/data/genotype_interpretations.parquet`
- `public/data/variant_keys.parquet`
- `public/data/source_summary.parquet`
- `public/data/index_summary.parquet`
- `public/data/size_report.tsv`
- `public/data/manifest.json`

See `docs/SIZE_TESTS.md` for measured ClinVar/GWAS bundle sizes, index overhead,
ZSTD compression-level tests, `parquet-linter` output, and Parquet row group
trade-offs.

## Committed annotation asset sizes

```{r asset-sizes, echo=FALSE, results='asis'}
size_path <- file.path("public", "data", "size_report.tsv")
if (file.exists(size_path)) {
  sizes <- utils::read.delim(size_path, check.names = FALSE)
  knitr::kable(sizes, align = c("l", "r", "r"))
} else {
  cat("`public/data/size_report.tsv` has not been generated yet.\n")
}
```

## Demo and test inputs

The repository includes small synthetic input files for testing both supported
input paths and both builds represented in the committed assets:

- `public/demo/example_23andme.txt`
- `public/demo/example_23andme_grch37.txt`
- `public/demo/example_23andme_grch38.txt`
- `public/demo/example_grch37.vcf`
- `public/demo/example_grch37.bcf`
- `public/demo/example_grch38.vcf`
- `public/demo/example_grch38.bcf`

## Repository layout

```text
index.html                         static app shell
src/duckgenesnap.js                browser webR/DuckDB/Rduckhts runtime
public/data/*.parquet              sorted annotation and metadata assets
public/demo/*                      synthetic chip/VCF/BCF test inputs
data/seed/*.tsv                    GRCh37 seed curation source
scripts/build_assets.R             optparse asset builder/injector
docs/                              design, schema, size notes
```

## Matching and storage policy

Core ingestion does **not** depend on rsID identity. Uploaded rows are matched by:

```text
analysis_build + chrom_norm + pos
```

`source_id` may contain rsIDs or source-specific identifiers, but it is display
metadata only. `variant_keys` is kept as an auxiliary allele-specific table for
VCF/BCF QA, detail panels, and future exact allele refinement.

Do not lift or mutate VariantKeys directly. For allele-aware paths:

```text
source chrom/pos/ref/alt
  -> optional normalization/liftover
  -> recompute variantkey(chrom, pos, ref, alt)
  -> optional exact-key refinement
```

## Upstream attribution and frontend query credit

DuckGeneSnap is an independent static-site/DuckDB/Rduckhts implementation
inspired by the upstream GeneSnap project by syao13:
<https://github.com/syao13/GeneSnap>. DuckGeneSnap keeps the user-facing idea
of parsing raw genetic data and annotating clinically significant variants, but
moves the analysis into a backend-free browser runtime.

The planned variant detail panels also take inspiration from the lightweight
frontend-only query orchestration in Sasha Gusev's GWAS Lookup:
<https://github.com/sashagusev/gwas_lookup>. DuckGeneSnap keeps that spirit
(client-side modules, URL-stateable variant queries, parallel enrichment panels)
but uses local DuckDB/Rduckhts assets first and calls external APIs only for
on-demand detail enrichment.

## Related projects

- Upstream GeneSnap: <https://github.com/syao13/GeneSnap>
- DuckBedQC: <https://github.com/sounkou-bioinfo/DuckBedQC>
- GWAS Lookup frontend inspiration: <https://github.com/sashagusev/gwas_lookup>
- Rduckhts / duckhts: <https://github.com/RGenomicsETL/duckhts>
